A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2024-05-07 tidyindex.qmd
small fixes

```{r tbl-gggi-weights}
#| tbl-cap: "Weights for the two applications of dimension reduction to compute the Global Gender Gap Index. V-wgt is used to compute four new variables from the original 14. These are then equally combined to get the final index value."
#| tbl-cap-location: bottom
# table cap location is not working: https://github.com/quarto-dev/quarto-cli/issues/6417
library(kableExtra)
gggi_weights %>%  
   mutate(variable = str_replace_all(variable, "_", " ") %>%
            str_to_sentence()) %>% 
  select(c(1, 5, 2, 6, 7)) %>% 
  mutate(dimension = c("Economy", rep("", 4), 
                       "Education", rep("" , 3), 
                       "Health", "", 
                       "Politics", rep("", 2)),
         dim_weight = c(0.25, rep("", 4),
                        0.25, rep("", 3),
                        0.25, "",
                        0.25, rep("", 2))) %>% 
  knitr::kable(
    digits = 3, booktabs = TRUE, 
    col.names = c("Variable", "V-wgt", "Dimension", "D-wgt", "wgt"),
    linesep = c("", "", "", "", "\\hline",
                "", "", "", "\\hline",
                "", "\\hline",
                "", "", "\\hline"),
      align = c("l", "r", "l", "r", "r")
    ) %>%
  kable_styling(font_size = 11) %>%
  row_spec(0, bold = TRUE) %>% 
  column_spec(1, width = "17.5em") %>% 
  column_spec(3, width = "3.5em") %>% 
  column_spec(c(2, 4), width = "4em") %>% 
  column_spec(5, width = "3em")
```

>>>>

```{r tbl-gggi-weights}
#| tbl-cap: "Weights for the two applications of dimension reduction to compute the Global Gender Gap Index. V-wgt is used to compute four new variables from the original 14. These are then equally combined to get the final index value."
#| tbl-cap-location: bottom
# table cap location is not working: https://github.com/quarto-dev/quarto-cli/issues/6417
library(kableExtra)
gggi_weights %>%  
   mutate(variable = str_replace_all(variable, "_", " ") %>%
            str_to_sentence()) %>% 
  select(c(1, 5, 2, 6, 7)) %>% 
  mutate(dimension = c("Economy", rep("", 4), 
                       "Education", rep("" , 3), 
                       "Health", "", 
                       "Politics", rep("", 2)),
         dim_weight = c(0.25, rep("", 4),
                        0.25, rep("", 3),
                        0.25, "",
                        0.25, rep("", 2))) %>% 
  knitr::kable(
    digits = 3, booktabs = TRUE, 
    col.names = c("Variable", "V-wgt", "Dimension", "D-wgt", "weight"),
    linesep = c("", "", "", "", "\\hline",
                "", "", "", "\\hline",
                "", "\\hline",
                "", "", "\\hline"),
      align = c("l", "r", "l", "r", "r")
    ) %>%
  kable_styling(font_size = 11) %>%
  row_spec(0, bold = TRUE) %>% 
  column_spec(1, width = "17.5em") %>% 
  column_spec(3, width = "3.5em") %>% 
  column_spec(c(2, 4), width = "4em") %>% 
  column_spec(5, width = "3em")
```

### 2024-05-07 tidyindex.tex
small fixes

The problem is that every index is developed in its own unique way, by
different researchers or organizations, and often indexes designed for
the same purpose cannot easily be compared. This echoes an issue raised
in Donoho (2017), that different data analysts might arrive at different
conclusions despite using the same data. This is especially pertinent to
index use which affects important decisions such as in natural disaster
prevention, economic interventions, resource allocation or human
development. It is primarily due to a lack of standards in data analysis
workflow. Donoho (2017) called for research on structuring a unified
workflow to address methodological variation across studies in data
science. Current practices also violate statistical principles, where
quantifying and understanding uncertainty are essential to deciding on a
best measure or metric, for example by incorporating bootstrap
confidence intervals (Efron 1979). There has been considerable research
in tidying up routine data analyses (Wickham 2014, 2011; Kuhn and Silge
2022; Wang, Cook, and Hyndman 2020; Zhang et al. 2024) that ``turn ideas
into software quickly and faithfully'', as envisioned by Chambers
(1998). Index development and use needs tidying.

@@

\caption{\label{fig-pipeline-steps}Diagram of pipeline modules for index
construction. The highlighted path illustrates one possible construction
using the dimension reduction and simplification module.}

@@

In what follows when geographic or temporal components of the
\(x_j(s,t)\) process are fixed we will be using suffix notation. For
example, \(x_{sj}(t)\) represents the data for a fixed location \(s\) as
a function of time \(t\). While \(x_{tj}(s)\) denotes the spatial
varying process for a fixed \(t\). An overview of the notation for
pipeline input, operation, and output is presented in
Table~\ref{tbl-notation}.

@@

The temporal processing module takes as input argument a single variable
\(x_{sj}(t)\) at location \(s\) as a function of time. In this step the
original time series can be transformed or summarized into a new one via
time aggregation, the transformation is represented by the function
\(f\), \(x^{\text{Temp}}_{sj}(t^\prime) = f[x_{sj}(t)]\) where
\(t^\prime\) refers to the new temporal resolution after aggregation. An
example of temporal processing done in the computation of the
Standardized Precipitation Index (SPI) (McKee et al. 1993), consists of
summing the monthly precipitation series over a rolling time window of
size \(k\). That is also known as the time scale. For SPI, the choice of
the time scale \(k\) is used to control the accumulation period for the
water deficit, enabling the assessment of drought severity across
various types (meteorological, agricultural, and hydrological).

@@

\caption{\label{fig-scale-var-trans-compare}Comparison of the module
scaling (green) and variable transformation (orange). While both modules
change the variable range, scaling maintains the same distributional
shape, which is not the case with variable transformation.}

@@

Although the scaling might be considered to be a transformation, we have
elected to make it a separate module because it is neater.
Figure~\ref{fig-scale-var-trans-compare} shows that scaling simply
changes the numbers in the data not the shape of a variable, while
transformation will most likely change the shape, and is usually
non-linear.

@@

\toprule
\textbf{Variable} & \textbf{V-wgt} & \textbf{Dimension} & \textbf{D-wgt} & \textbf{wgt}\\
\midrule
Labour force participation & 0.199 & Economy & 0.25 & 0.050\\
Wage equality for similar work & 0.310 &  &  & 0.078\\
Estimated earned income & 0.221 &  &  & 0.055\\
Legislators senior officials and managers & 0.149 &  &  & 0.037\\
Professional and technical workers & 0.121 &  &  & 0.030\\
\addlinespace
Literacy rate & 0.191 & Education & 0.25 & 0.048\\
Enrolment in primary education & 0.459 &  &  & 0.115\\
Enrolment in secondary education & 0.230 &  &  & 0.058\\
Enrolment in tertiary education & 0.121 &  &  & 0.030\\
Sex ratio at birth & 0.693 & Health & 0.25 & 0.173\\
\addlinespace
Healthy life expectancy & 0.307 &  &  & 0.077\\
Women in parliament & 0.310 & Politics & 0.25 & 0.078\\
Women in ministerial positions & 0.247 &  &  & 0.062\\
Years with female head of state & 0.443 &  &  & 0.111\\
\bottomrule

@@

Figure~\ref{fig-idx-tour} illustrates doing sensitivity analysis for
GGGI, for a subset of 16 countries. It presents 6 frames selected from
an animation where the weight on the politics dimension is gradually
increased, while other dimensions (economy, education, health) decrease
correspondingly. Frame 12 present the original index where all the four
dimensions receive equal weight. The index values are sorted from
highest to lowest, with the Nordic countries (Iceland, Norway, and
Finland) and New Zealand leading the rankings. The index values are
between 0 and 1, and indicate proportional difference between men and
women, with a value of 0.8 indicating women are 80\% of the way to
equality of these measures. There is a gap in values between these
countries and the middle group (Brazil, Panama, Poland, Bangladesh,
Kazakhstan, Armenia, and Slovakia), and another big drop to the next
group (Pakistan, Iran, Algeria, and Chad). Afghanistan lags much further
behind.

@@

Frames 1 and 6 show linear combinations where politics contributes less
than the original. It is interesting to note that the gap between the
Nordic countries and the middle countries dissipates, indicating that
this component was one reason for the relatively higher GGGI values of
these countries. Also interesting is the large drop in value for
Bangladesh. Frames 18, 24, 29 show linear combinations where politics
contributes more than the original. The most notable feature is that
Bangladesh retains it's high index value whereas the other middle group
countries decline, indicating that the politics score is a major
component for Bangladesh's index value.

Ideally, an index should be robust against minor changes in its
construction components. This is not the case with GGGI, where small
changes to one component lead to fairly large change in the index. The
modular pipeline framework for computing the index makes it easy to
conduct this type of sensitivity analysis, where one or more components
are perturbed and the index re-calculated. One aspect of the GGGI not
well-described in the Global Gender Gap Report is handling of missing
values that are present in the initial variables for many countries,
something that is common for this type of data. This could also be made
more transparent with the dimension reduction module, by specifying an
imputation method or providing warnings about missing values.

@@

The confidence interval can then be calculated using the quantile method
from the bootstrap samples. Figure~\ref{fig-conf-interval} presents the
80\% and 95\% confidence interval for the Texas post office station, in
Queensland, Australia. From the start of 2019 to 2020, the majority of
the confidence intervals lie below the extreme drought line (SPI = -2),
suggesting a high level of certainty that the Texas post office is
suffering from a drastic drought. Also close to the extreme drought line
is the period 2003-2004, which corresponds to the millennium drought.
These relatively wide confidence intervals, as well as during the
excessive precipitation events in 1996-1998 and 1999-2000, suggest a
high variation of the gamma parameters estimated from the bootstrap
samples and its difficulty to accurately quantify the drought and flood
severity in extreme events.

@@

\caption{\label{fig-conf-interval}80\% and 95\% confidence interval of
the Standardized Precipitation Index (SPI-24) for the Texas post office
station, in Queensland, Australia. A bootstrap sample of 100 is taken
from the aggregated precipitation series to estimate gamma parameters
and to calculate the index. The dashed line at SPI = -2 represents an
extreme drought as defined by the SPI. Most parts of the confidence
intervals from 2019 to 2020 sit below the extreme drought line and are
relatively wide compared to other time periods. This suggests that while
it is certain that the Texas post office is suffering from a drastic
drought, there is considerable uncertainty in quantifying its severity,
given the extremity of the event.}

@@

\bibitem[\citeproctext]{ref-chambers1998programming}
Chambers, John M. 1998. \emph{Programming with Data: A Guide to the s
Language}. Berlin, Heidelberg: Springer-Verlag.

@@

\bibitem[\citeproctext]{ref-tidymodels}
Kuhn, Max, and Julia Silge. 2022. \emph{Tidy Modeling with r}. "
O'Reilly Media, Inc.".

@@

\bibitem[\citeproctext]{ref-zhang2024cubble}
Zhang, H. Sherry, Dianne Cook, Ursula Laa, Nicolas Langrené, and
Patricia Menéndez. 2024. {``Cubble: An r Package for Organizing and
Wrangling Multivariate Spatio-Temporal Data.''}
\url{https://arxiv.org/abs/2205.00259}.

>>>>

The problem is that every index is developed in its own unique way, by
different researchers or organizations, and often indexes designed for
the same purpose cannot easily be compared. This echoes an issue raised
in Donoho (2017), that different data analysts might arrive at different
conclusions despite using the same data. This is especially pertinent to
index use which affects important decisions such as in natural disaster
prevention, economic interventions, resource allocation or human
development. It is primarily due to a lack of standards in data analysis
workflow. Donoho (2017) called for research on structuring a unified
workflow to address methodological variation across studies in data
science. Current practices also violate statistical principles, where
quantifying and understanding uncertainty are essential to deciding on a
best measure or metric, for example by incorporating bootstrap
confidence intervals (Efron 1979). There has been considerable research
in tidying up routine data analyses (Wickham 2014, 2011; Kuhn and Silge
2022; Wang, Cook, and Hyndman 2020; Zhang et al. to appear) that ``turn
ideas into software quickly and faithfully'', as envisioned by Chambers
(1998). Index development and use needs tidying.

@@

\caption{\label{fig-pipeline-steps}Diagram of pipeline modules for index
construction. The highlighted path illustrates one possible construction
using the dimension reduction and simplification modules.}

@@

In what follows when geographic or temporal components of the
\(x_j(s,t)\) process are fixed we will be using suffix notation. For
example, \(x_{sj}(t)\) represents the data for a fixed location \(s\) as
a function of time \(t\), while \(x_{tj}(s)\) denotes the spatial
varying process for a fixed \(t\). An overview of the notation for
pipeline input, operation, and output is presented in
Table~\ref{tbl-notation}.

@@

The temporal processing module takes as input argument a single variable
\(x_{sj}(t)\) at location \(s\) as a function of time. In this step, the
original time series can be transformed or summarized into a new one via
time aggregation. The transformation is represented by the function
\(f\), \(x^{\text{Temp}}_{sj}(t^\prime) = f[x_{sj}(t)]\) where
\(t^\prime\) refers to the new temporal resolution after aggregation. An
example of temporal processing done in the computation of the
Standardized Precipitation Index (SPI) (McKee et al. 1993), consists of
summing the monthly precipitation series over a rolling time window of
size \(k\). That is also known as the time scale. For SPI, the choice of
the time scale \(k\) is used to control the accumulation period for the
water deficit, enabling the assessment of drought severity across
various types (meteorological, agricultural, and hydrological).

@@

\caption{\label{fig-scale-var-trans-compare}Comparison of the scaling
(green) and variable transformation (orange) modules. While both modules
change the variable range, scaling maintains the same distributional
shape, which is not the case with variable transformation.}

@@

Although the scaling might be considered to be a transformation, we have
elected to make it a separate module because it is neater.
Figure~\ref{fig-scale-var-trans-compare} shows that scaling simply
changes the numbers in the data but not the shape of a variable, while
transformation will most likely change the shape, as it is usually
non-linear.

@@

\toprule
\textbf{Variable} & \textbf{V-wgt} & \textbf{Dimension} & \textbf{D-wgt} & \textbf{weight}\\
\midrule
Labour force participation & 0.199 & Economy & 0.25 & 0.050\\
Wage equality for similar work & 0.310 &  &  & 0.078\\
Estimated earned income & 0.221 &  &  & 0.055\\
Legislators senior officials and managers & 0.149 &  &  & 0.037\\
Professional and technical workers & 0.121 &  &  & 0.030\\
\addlinespace
Literacy rate & 0.191 & Education & 0.25 & 0.048\\
Enrolment in primary education & 0.459 &  &  & 0.115\\
Enrolment in secondary education & 0.230 &  &  & 0.058\\
Enrolment in tertiary education & 0.121 &  &  & 0.030\\
Sex ratio at birth & 0.693 & Health & 0.25 & 0.173\\
\addlinespace
Healthy life expectancy & 0.307 &  &  & 0.077\\
Women in parliament & 0.310 & Politics & 0.25 & 0.078\\
Women in ministerial positions & 0.247 &  &  & 0.062\\
Years with female head of state & 0.443 &  &  & 0.111\\
\bottomrule

@@

Figure~\ref{fig-idx-tour} illustrates doing sensitivity analysis for
GGGI, for a subset of 16 countries. It presents 6 frames selected from
an animation where the weight on the politics dimension is gradually
increased, while other dimensions (economy, education, health) decrease
correspondingly. Frame 12 presents the original index where all the four
dimensions receive equal weight. The index values are sorted from
highest to lowest, with the Nordic countries (Iceland, Norway, and
Finland) and New Zealand leading the rankings. The index values are
between 0 and 1, and indicate proportional difference between men and
women, with a value of 0.8 indicating women are 80\% of the way to
equality of these measures. There is a gap in values between these
countries and the middle group (Brazil, Panama, Poland, Bangladesh,
Kazakhstan, Armenia, and Slovakia), and another big drop to the next
group (Pakistan, Iran, Algeria, and Chad). Afghanistan lags much further
behind.

@@

Frames 1 and 6 show linear combinations where politics contributes less
than the original. It is interesting to note that the gap between the
Nordic countries and the middle countries dissipates, indicating that
this component was one reason for the relatively higher GGGI values of
these countries. Also interesting is the large drop in value for
Bangladesh. Frames 18, 24, 29 show linear combinations where politics
contributes more than the original. The most notable feature is that
Bangladesh retains its high index value whereas the other middle group
countries decline, indicating that the politics score is a major
component for Bangladesh's index value.

Ideally, an index should be robust against minor changes in its
construction components. This is not the case with GGGI, where small
changes to one component lead to fairly large change in the index. The
modular pipeline framework for computing the index makes it easy to
conduct this type of sensitivity analysis, where one or more components
are perturbed and the index recalculated. One aspect of the GGGI not
well-described in the Global Gender Gap Report is the handling of
missing values that are present in the initial variables for many
countries, something that is common for this type of data. This could
also be made more transparent with the dimension reduction module, by
specifying an imputation method or providing warnings about missing
values.

@@

The confidence interval can then be calculated using the quantile method
from the bootstrap samples. Figure~\ref{fig-conf-interval} presents the
80\% and 95\% confidence intervals for the Texas post office station, in
Queensland, Australia. From the start of 2019 to 2020, the majority of
the confidence intervals lie below the extreme drought line (SPI = -2),
suggesting a high level of certainty that the Texas post office is
suffering from a drastic drought. Also close to the extreme drought line
is the period 2003-2004, which corresponds to the millennium drought.
These relatively wide confidence intervals, as well as during the
excessive precipitation events in 1996-1998 and 1999-2000, suggest a
high variation of the gamma parameters estimated from the bootstrap
samples and its difficulty to accurately quantify the drought and flood
severity in extreme events.

@@

\caption{\label{fig-conf-interval}80\% and 95\% confidence intervals of
the Standardized Precipitation Index (SPI-24) for the Texas post office
station, in Queensland, Australia. A bootstrap sample of 100 is taken
from the aggregated precipitation series to estimate gamma parameters
and to calculate the index. The dashed line at SPI = -2 represents an
extreme drought as defined by the SPI. Most parts of the confidence
intervals from 2019 to 2020 sit below the extreme drought line and are
relatively wide compared to other time periods. This suggests that while
it is certain that the Texas post office is suffering from a drastic
drought, there is considerable uncertainty in quantifying its severity,
given the extremity of the event.}

@@

\bibitem[\citeproctext]{ref-chambers1998programming}
Chambers, John M. 1998. \emph{Programming with Data: A Guide to the {S}
Language}. Berlin, Heidelberg: Springer-Verlag.

@@

\bibitem[\citeproctext]{ref-tidymodels}
Kuhn, Max, and Julia Silge. 2022. \emph{Tidy Modeling with {R}}. "
O'Reilly Media, Inc.".

@@

\bibitem[\citeproctext]{ref-zhang2024cubble}
Zhang, H. Sherry, Dianne Cook, Ursula Laa, Nicolas Langrené, and
Patricia Menéndez. to appear. {``Cubble: An {R} Package for Organizing
and Wrangling Multivariate Spatio-Temporal Data.''} \emph{Journal of
Statistical Software}, to appear.
